---
title: "London vehicle travel time STDM: data preparation"
author: "Mark Ruddy"
date: "27 March 2015"
output: html_document
---

# Description

EDA for space-time aurocorrelation analysis of network data for London motor vehicle travel time analysis.


# Imports

## Packages
```{r Package imports}
library(gdata) # for xls import
library(ggplot2)
library(dplyr)
library(tidyr)
#library(zoo)
library(xts)
#library(lubridate)
library(chron)
```


# Space-time autocorrelation analysis

## Road network data

Network aggregate motor vehicle travel times in seconds per metre from 22 road links.

Data taken between Monday 24th May 2010 and Friday 5th November 2010 (23 weeks and five days).

The motor vehicle travel time data comes as a directed network.

Spatial weights matrices for the travel data are a binary scheme row standardised. Standardisation is done to equally allocate the contribution of neighbouring nodes to the node of interest ([@cheng2012spatio p396] for details).

## Data preparation

Read in network aggregate travel time data (unit journey time - ujt). 
```{r Read ujt data}
ujt <- read.csv("./data/ujt.csv")
```

Read in spatial weights. Need to open .xls file and copy/paste weight matrices into separate worksheets, then refer to individual worksheets during read.xls()
Get weights from .xls file and write each to a new .csv
```{r}
# w1 <- read.xls(xls = "./data/Weight_Matrices.xls", sheet="w1",header =T)
# write.csv(w1, "./data/weights1.csv", row.names=F)
# w2 <- read.xls(xls = "./data/Weight_Matrices.xls", sheet="w2",header =T)
# write.csv(w2, "./data/weights2.csv", row.names=F)
# w3 <- read.xls(xls = "./data/Weight_Matrices.xls", sheet="w3",header =T)
# write.csv(w3, "./data/weights3.csv", row.names=F)
```

```{r Read spatio-temporal weights}
w1 <- read.csv("./data/weights1.csv")
w2 <- read.csv("./data/weights2.csv")
w3 <- read.csv("./data/weights3.csv")
```


## Make date-time series for ujt observations
Now we need to add the dates and times when each observation was made. This is for data selection and plotting purposes.

Create data and times
```{r dates and times for observations}
# create time intervals for one day - 6:00 to 21:00
t <- as.character(times(72:252/288))
#d <- timeBasedSeq("2010-05-24/2010-11-04/d") Bug in timeBasedSeq gives repeated 31st Oct at Daylight Saving time
d <- seq.Date(as.Date("2010-05-24"),as.Date("2010-11-05"), 1)
obs.times <- rep(t, times=length(d))
obs.dates <- sort(as.Date(rep(d, times=length(t)), format="%d-%m-%y"))
datetimes<-strptime(paste(obs.dates,obs.times),format="%Y-%m-%d %H:%M", tz="GMT")
rm(t,d)
```

Create ujt data.frame with dates-times for plotting. Add dates to ujt with dates and times as factors and ujt as data.frame.

```{r Create ujt data.frame with days}
ujt.dt <- cbind(as.factor(obs.times), as.factor(obs.dates), ujt)
colnames(ujt.dt)[1:2] <- c("time", "date")

```

Add decimal time to ujt
```{r Add decimal time to ujt}
# add decimal times for plotting
# ref http://stackoverflow.com/questions/5186972/how-to-convert-time-mmss-to-decimal-form-in-r
obs.dec <- sapply(strsplit(obs.times,":"),
                      function(x) {
                        x <- as.numeric(x)
                        x[1]+x[2]/60
                      })

# combine with time as factor
ujt.dt <- cbind(obs.dec,ujt.dt)
colnames(ujt.dt)[1] <- "dtime"
# head(ujt.dt)

# clean up
rm(datetime, obs.dec, t,d, obs.times, obs.dates)
```



