---
title: "Outlier detection"
author: "Mark Ruddy"
date: "27 March 2015"
output: html_document
---

# Description

Outlier detection from ujt dataset.

Approach is initial visual inspection of individual journey times for each link.

Scatterplot of times per link throughout time series.

Try for one link first


## Imports
### Packages and functions.
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(grid) # for arrows
```

### Data

Spatial weights and unit journey time datasets.

```{r}
load("./data/STDM_base.RData")
```

## Outliers for link X2085

Plot link X1447 over whole time series:


```{r outlier examination for all links}
ujt.wed <- ujt.dt %>%
  filter(weekdays(as.Date(date)) %in% "Wednesday") %>%
  select(-X1384, -X2007, -X2140, -X2301)
  #View()

ujt.g <- ujt.wed %>% gather(link, ujt, -dtime:-date)

p <- ggplot(ujt.g, aes(x=dtime, y=ujt))

p + 
  theme(panel.grid.minor=element_blank()) +
  geom_point(alpha=0.5, size=0.8) +
  facet_wrap(~ link, ncol=3, scales="fixed") +
  scale_x_continuous(limits=c(6,21), breaks=6:21) + 
  labs(title="Unit journey times", x="Time of day", y="Unit Journey Time (seconds/metre)")
# Unit journey times for links (excluding links with historic data).

ggsave("./outputs/ujt_scatters.pdf", width=10, height=10)
# clean up
#rm(p, ujt.wed, ujt.g)
```

Examine a few links in detail for outlier
```{r outlier examination in X474}
u <- ujt.g %>% filter(link=="X474")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + geom_point(alpha=0.5, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date," ", u$time),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X474", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-05-26"], y=ujt[date=="2010-05-26"]), colour="grey10", size=0.5) + 
  geom_line(aes(x=dtime[date=="2010-07-07"], y=ujt[date=="2010-07-07"]), colour="grey10", size=0.5, linetype = 2) + 
  geom_line(aes(x=dtime[date=="2010-09-08"], y=ujt[date=="2010-09-08"]), colour="grey10", size=0.5, linetype = 4) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=8, y=max(u$ujt), label="2010-05-26", hjust=-0.1) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt)*.9, yend=max(u$ujt)*.9, linetype=2) +
  annotate(geom="text", x=8, y=max(u$ujt)*.9, label="2010-07-07", hjust=-0.1) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt)*.8, yend=max(u$ujt)*.8, linetype=4) +
  annotate(geom="text", x=8, y=max(u$ujt)*.8, label="2010-09-08", hjust=-0.1)

# clean up
rm(olimit, p)
```

Examine a few links in detail for outliers
```{r outlier examination in X425}
u <- ujt.g %>% filter(link=="X425")
olimit <- 8*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X425", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-05-26"], y=ujt[date=="2010-05-26"]), colour="grey10", size=0.5) + 
  geom_line(aes(x=dtime[date=="2010-07-21"], y=ujt[date=="2010-07-21"]), colour="grey10", size=0.5, linetype = 2) + 
  geom_line(aes(x=dtime[date=="2010-07-28"], y=ujt[date=="2010-07-28"]), colour="grey10", size=0.5, linetype = 4) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=19, y=max(u$ujt), label="2010-05-26", hjust=-0.1) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt)*.9, yend=max(u$ujt)*.9, linetype=2) +
  annotate(geom="text", x=19, y=max(u$ujt)*.9, label="2010-07-21", hjust=-0.1) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt)*.8, yend=max(u$ujt)*.8, linetype=4) +
  annotate(geom="text", x=19, y=max(u$ujt)*.8, label="2010-07-28", hjust=-0.1)

# clean up
rm(olimit, p)
```

Examine a few links in detail for outliers
```{r outlier examination in X1447}
u <- ujt.g %>% filter(link=="X1447")
olimit <- 8*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X1447", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-05-26"], y=ujt[date=="2010-05-26"]), colour="grey10", size=0.5) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=8, y=max(u$ujt), label="2010-05-26", hjust=-0.1)

# clean up
rm(olimit, p)
```

Examine a few links in detail for outliers
```{r outlier examination in X542}
u <- ujt.g %>% filter(link=="X524")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X524", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-06-09"], y=ujt[date=="2010-06-09"]), colour="grey10", size=0.5) + 
  geom_line(aes(x=dtime[date=="2010-06-16"], y=ujt[date=="2010-06-16"]), colour="grey10", size=0.5, linetype=2) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=8, y=max(u$ujt), label="2010-06-09", hjust=-0.1) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt)*.9, yend=max(u$ujt)*.9, linetype=2) +
  annotate(geom="text", x=8, y=max(u$ujt)*.9, label="2010-06-16", hjust=-0.1)

# clean up
rm(olimit, p)
```


Examine a few links in detail for outliers
```{r outlier examination in X1025}
u <- ujt.g %>% filter(link=="X1025")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + 
  geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X1025", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-06-09"], y=ujt[date=="2010-06-09"]), colour="grey10", size=0.5) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  annotate(geom="segment", x=6, xend=7.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=8, y=max(u$ujt), label="2010-06-09", hjust=-0.1)

# clean up
rm(olimit, p)
```

Examine a few links in detail for outliers
```{r outlier examination in X1592}
u <- ujt.g %>% filter(link=="X1592")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + 
  geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X1592", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-05-26"], y=ujt[date=="2010-05-26"]), colour="grey10", size=0.5) +
  geom_line(aes(x=dtime[date=="2010-07-21"], y=ujt[date=="2010-07-21"]), colour="grey10", size=0.5, linetype=2) + 
  geom_line(aes(x=dtime[date=="2010-10-20"], y=ujt[date=="2010-10-20"]), colour="grey10", size=0.5, linetype=4) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
    annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=19, y=max(u$ujt), label="2010-05-26", hjust=-0.1) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt)*.9, yend=max(u$ujt)*.9, linetype=2) +
  annotate(geom="text", x=19, y=max(u$ujt)*.9, label="2010-07-21", hjust=-0.1) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt)*.8, yend=max(u$ujt)*.8, linetype=4) +
  annotate(geom="text", x=19, y=max(u$ujt)*.8, label="2010-10-20", hjust=-0.1)

# clean up
rm(olimit, p)
```


Examine a few links in detail for outliers
```{r outlier examination in X2052}
u <- ujt.g %>% filter(link=="X2052")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + 
  geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X2052", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-05-26"], y=ujt[date=="2010-05-26"]), colour="grey10", size=0.5) +
  geom_line(aes(x=dtime[date=="2010-06-30"], y=ujt[date=="2010-06-30"]), colour="grey10", size=0.5, linetype=2) + 
  geom_line(aes(x=dtime[date=="2010-09-29"], y=ujt[date=="2010-09-29"]), colour="grey10", size=0.5, linetype=4) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt), yend=max(u$ujt)) +
  annotate(geom="text", x=19, y=max(u$ujt), label="2010-05-26", hjust=-0.1) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt)*.9, yend=max(u$ujt)*.9, linetype=2) +
  annotate(geom="text", x=19, y=max(u$ujt)*.9, label="2010-06-30", hjust=-0.1) +
  annotate(geom="segment", x=17, xend=18.5, y=max(u$ujt)*.8, yend=max(u$ujt)*.8, linetype=4) +
  annotate(geom="text", x=19, y=max(u$ujt)*.8, label="2010-09-29", hjust=-0.1)

# clean up
rm(olimit, p)
```

Examine a few links in detail for outliers
```{r outlier examination in X2079}
u <- ujt.g %>% filter(link=="X2079")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + 
  geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X2079", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-05-26"], y=ujt[date=="2010-05-26"]), colour="grey10", size=0.5) + 
  geom_line(aes(x=dtime[date=="2010-07-28"], y=ujt[date=="2010-07-28"]), colour="grey10", size=0.5, linetype=3) +
  geom_line(aes(x=dtime[date=="2010-09-01"], y=ujt[date=="2010-09-01"]), colour="grey10", size=0.5, linetype=5) + 
  #geom_line(aes(x=dtime[date=="2010-09-15"], y=ujt[date=="2010-09-15"]), colour="grey10", size=0.5, linetype=5) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) 

# clean up
rm(olimit, p)
```


Examine a few links in detail for outliers
```{r outlier examination in X2085}
u <- ujt.g %>% filter(link=="X2085")
olimit <- 7*IQR(u$ujt)
p <- ggplot(u, aes(x=dtime, y=ujt))
p + 
  geom_point(alpha=0.4, colour="green4", size=2) +
  theme(panel.grid.minor=element_blank()) +
  geom_vline(xintercept=10, colour="grey50", size=1) +
  geom_vline(xintercept=16, colour="grey50", size=1) +
  annotate("text", x=8, y= -0.05, label= "AMP") +
  annotate("text", x=13, y=-0.05, label= "IP") +
  annotate("text", x=18.5, y=-0.05, label= "PMP") +
  #geom_text(aes(label=ifelse((ujt>olimit), paste(date),"")), size=3, hjust=1.1) + 
  labs(title="Unit journey times for link X2085", x="Time of day", y="Unit Journey Time (seconds/metre)") +
  geom_line(aes(x=dtime[date=="2010-07-21"], y=ujt[date=="2010-07-21"]), colour="grey10", size=0.5) +
  #geom_line(aes(x=dtime[date=="2010-06-30"], y=ujt[date=="2010-06-30"]), colour="grey10", size=0.5, linetype=3) + 
  #geom_line(aes(x=dtime[date=="2010-09-29"], y=ujt[date=="2010-09-29"]), colour="grey10", size=0.5, linetype=5) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) 

# clean up
rm(olimit, p)
```










