---
title: "London vehicle travel time STDM: EDA"
author: "Mark Ruddy"
date: "24 March 2015"
output: html_document
---

# Description

Data preparation for space-time aurocorrelation analysis of network data for London motor vehicle travel time analysis.

# Imports
## Packages
```{r Package imports}
library(gdata) # for xls import
library(ggplot2)
library(dplyr)
library(tidyr)
#library(zoo)
library(xts)
#library(lubridate)
library(chron)
```

## Datasets
```{r}
load("./data/STDM_base.RData")
```


# Time series averages from single days

We want to plot averages of time series from just one day.

## Picking weekdays

Tuesday, Wednesdays or Thursdays are assumed to be generally representative of traffic flows. Aim: to extract time series only for midweek days and plot for specific links to determine whether all midweek days are comparable and to contrast with Mondays, Fridays and Weekends.

### Plot Tuesdays 

For a selection of links (X425, X2140, X432, X1419) or X1616, X463, X1593, X2324 which correspond with Cheng et al 2011.
1. Filter day
2. Group by time
3. Find means for links
```{r Plot Tuesdays, fig.width=12}
ujt.tues <- ujt.dt %>%
  group_by(dtime) %>%
  filter(weekdays(as.Date(date)) %in% "Tuesday") %>%
  summarise_each(funs(mean)) %>%
  select(dtime, X1616, X463, X1593, X2324)
# checked for number of inputs to each mean by using summarise_each(func(mean,n())

# gather data for plotting
ujt.tues.g <- ujt.tues %>% gather(link, ujt_mean, -dtime)

# plot mean speed for 
# 1 metre/second = 2.23694 miles per hour ref google (1/ujt_mean*2.23694)
ggplot(ujt.tues.g, aes(x=dtime, y=1/ujt_mean*2.23694, colour=link, group=link)) + 
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  scale_y_continuous(limits=c(0,15)) +
  labs(title="Variation in mean traffic speed for Tuesdays", x="Time of day (hour)", y="Mean journey speed (miles/hour)") +
  geom_line()

rm(ujt.tues.g, ujt.tues)
```

### Plot Tuesdays, Wednesdays, and Thursdays 
To compare traffic flows between midweek days.

For links c("X1616", "X2085", "X432", "X1419") in 4 different facets across . 
```{r Plot some links in mid-week, fig.width=12, fig.height=10}
# add weekdays and get means for each time to give ujt df for links we're interested in
ujt.wkd <- ujt.dt %>%
  mutate(day = weekdays(as.Date(date))) %>%
  group_by(dtime, day) %>%
  summarise_each(funs(mean)) %>%
  select(dtime, day, X425, X2085, X432, X1419, X463, X1593, X1447)

# tidy data
ujt.wkd.g <- ujt.wkd %>% gather(link, mean, -dtime:-day)

p <- ggplot(ujt.wkd.g, aes(x=dtime, y=mean, colour=day))

p + geom_line() + 
  facet_wrap(~ link, nrow=3, scales="free_y") +
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  #scale_y_continuous(limits=c(0.05,0.35)) +
  labs(title="Average unit journey time profiles", x="Time of day (hour)", y="Mean unit journey time (seconds/metre)")

rm(p,ujt.wkd, ujt.wkd.g)
```

Plot of historical Link data
```{r}
# add weekdays and get means for each time to give ujt df for links we're interested in
ujt.wkd.hist <- ujt.dt %>%
  mutate(day = weekdays(as.Date(date))) %>%
  group_by(dtime, day) %>%
  summarise_each(funs(mean)) %>%
  select(dtime, day, X1384, X2007, X2140, X2301)

ujt.wkd.hist.g <- ujt.wkd.hist %>% gather(link, mean, -dtime:-day)

p <- ggplot(ujt.wkd.hist.g, aes(x=dtime, y=mean, colour=day))

p + geom_line() + 
  facet_wrap(~ link, nrow=3, scales="free_y") +
  scale_x_continuous(limits=c(6,21), breaks=6:21) +
  #scale_y_continuous(limits=c(0.05,0.35)) +
  labs(title="Average unit journey time profiles for historical links", x="Time of day (hour)", y="Mean unit journey time (seconds/metre)")

# clean up
rm(p, ujt.wkd.hist, ujt.wkd.hist.g)
```





