---
title: "Space-time traffic SVMs for whole day"
author: "Mark Ruddy"
date: "17 April 2015"
output: html_document
---


# Tuning SVM model of traffic data from whole day
1. Import data
2. Train models using 1st order spatial weights
3. Try different lengths of hsitorical data for regression (m=c(3, 6, 12): 15, 30 and 60 minutes)
4. Choose two links for model assessment (X448, X1593)
5. Train with 80% training data

## Setup: data, functions and packages

```{r Setup: data, functions and packages}
library(kernlab)
library(ggplot2)
library(grid)
library(dplyr)
library(tidyr)

source("./functions/Kernel_methods_lib.R")
source("./functions/trainModel.R") # train models with mutiple parameters
source("./functions/trainTraffic.R") # train models with mutiple parameters from multiple datasets
source("./functions/Kernel_methods_lib.R")
  
# datasets including spatial weights and Wednesday UJTs
load("./data/STDM_base.RData")

# rmse function
rmse <- function(obs, pred, na.rm=TRUE) {
  sqrt(mean((obs-pred)^2, na.rm=na.rm))
  }

# Wednesday UJTs
t <- ujt.repl.hist
```


Prepare data
```{r Prepare weight matrices}
# weight matrices as.matrix
w1.m <- as.matrix(w1[,-1])
w2.m <- as.matrix(w2[,-1])
w3.m <- as.matrix(w3[,-1])

```


## Train UJT for links X448 and X1593

NB:embed time series
# m: historical period over which prediction is based
# col: data to be modelled
# W: weights matrix (the spatial neighbours that affect data to be modelled)
# ii: whether to use 'col' data in modelling process
stseries_m3 <- st_embed(data=select(t,-dtime:-date), m=3, col=1, W=w1.m, ii=TRUE)
# gives matrix X of dimensions nrow(data)*(m*(number of spatial neighbours if ii=TRUE)+1)
# gives vector y of 'col' variables of length (m+1)*nrow(data)
# eg if m=3, and col="X425", 1st order W specifies neighbours with X2140 and X1592. This gives a nrow(data)*9 matrix X.

```{r Train UJT}
# parmaters
C <- 2^rep(-5:4) # 
sigma <- 2^rep(-9:2) # 
epsilon <-c(0.01, 0.1, 0.5) # 


# split at 3620 is 20 days of Wednesday observations (20 days * 181 5min intervals) 
v.day.m3 <- trainTraffic(ujt = select(t, -dtime:-date), weights = w1.m, link= c(3, 12), m = 3, C = C, epsilon = epsilon, sigma = sigma, split = 3620)

v.day.m6 <- trainTraffic(ujt = select(t, -dtime:-date), weights = w1.m, link = c(3, 12), m = 6, C = C, epsilon = epsilon, sigma = sigma, split = 3620)

v.day.m12 <- trainTraffic(ujt = select(t, -dtime:-date), weights = w1.m, link = c(3, 12), m = 12, C = C, epsilon = epsilon, sigma = sigma, split = 3620)

```

## Plot errors for X448
Combine x448 training for all m
```{r Combine x448 training for all m}
v3 <- v.day.m3$X448$error_mat
v6 <- v.day.m6$X448$error_mat
v12 <- v.day.m12$X448$error_mat

v448 <- rbind(v3, v6, v12)
v448 <- cbind(v448, c(rep(3,nrow(v3)), rep(6,nrow(v6)), rep(12,nrow(v12))))
colnames(v448)[8] <- "m" 
v448 <- as.data.frame(v448)

rm(v3, v6, v12)
```

Plot multi-facet of X448 for training error ~ different epsilon and m
```{r Plot multi-facet of X448 training}

p <- ggplot(v448, aes(x=factor(Sigma), y=factor(C), fill=log(TrainingError)))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR training error - link X448 with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="log Training RMSE"), low="#000066", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
#ggsave(filename = "./outputs/TrainError_X448.pdf",width = 15, height = 9)
ggsave(filename = "./outputs/TrainError_X448.png",width = 15, height = 9)

```


Plot multi-facet of X448 for testing error - different epsilon and m
```{r Plot multi-facet of X448 testing}

p <- ggplot(v448, aes(x=factor(Sigma), y=factor(C), fill=log(RMSE)))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR test error - link X448 with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="log Forecast RMSE"), low="darkred", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
#ggsave(filename = "./outputs/TestError_X448.pdf",width = 15, height = 9)
ggsave(filename = "./outputs/TestError_X448.png",width = 15, height = 9)

```



## Plot errors for X1593
Combine X1593 training for all m
```{r Combine X1593 training for all m}
v3 <- v.day.m3$X1593$error_mat
v6 <- v.day.m6$X1593$error_mat
v12 <- v.day.m12$X1593$error_mat

v1593 <- rbind(v3, v6, v12)
v1593 <- cbind(v1593, c(rep(3,nrow(v3)), rep(6,nrow(v6)), rep(12,nrow(v12))))
colnames(v1593)[8] <- "m" 
v1593 <- as.data.frame(v1593)

rm(v3, v6, v12)
```

Plot multi-facet of X1593 for training error ~ different epsilon and m
```{r Plot multi-facet of X1593 training}

p <- ggplot(v1593, aes(x=factor(Sigma), y=factor(C), fill=log(TrainingError)))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR training error - link X1593 with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="log Training RMSE"), low="#000066", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
#ggsave(filename = "./outputs/TrainError_X1593.pdf",width = 15, height = 9)
ggsave(filename = "./outputs/TrainError_X1593.png",width = 15, height = 9)

```

Plot multi-facet of X1593 for testing error ~ different epsilon and m
```{r Plot multi-facet of X1593 testing}

p <- ggplot(v1593, aes(x=factor(Sigma), y=factor(C), fill=log(RMSE)))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR test error - link X1593 with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="log Forecast RMSE"), low="darkred", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
#ggsave(filename = "./outputs/TestError_X1593.pdf",width = 15, height = 9)
ggsave(filename = "./outputs/TestError_X1593.png",width = 15, height = 9)

```

Detailed data checking
```{r Detailed data checking}
# RMSE
v448 %>% group_by(Epsilon, m) %>% summarise(min(TrainingError), min(RMSE)) %>% View()

v1593 %>% group_by(Epsilon, m) %>% summarise(min(TrainingError), min(RMSE)) %>% View()

# Running times

v448 %>% group_by(C, m) %>% slice(which.min(time)) %>%  View()

v1593 %>% group_by(C, m) %>% slice(which.min(time)) %>%  View()

```


Plot models 
```{r plot best model for 448 and 1593}

# Final model parameter space, estimated from model training above
# Compromise between potentially overfitting model and allowing generalisation
# C <- 2^2
# epsilon <- 0.10
# sigma <- 2^-4
# m <- 6

# source("./functions/trainTraffic.R") # train models with mutiple parameters from multiple datasets
# split at 3620 is 20 days of Wednesday observations (20 days * 181 5min intervals) 


split=3620
C <- 2^2
epsilon <- 0.10
sigma <- 2^-4
m <- 6
t.fcasts <- select(t, -dtime:-date)

ptm <- proc.time()
fcasts <- trainTraffic(ujt = t.fcasts, weights = w1.m, link = c(1:dim(t.fcasts)[2]), m = m, C = C, epsilon = epsilon, sigma = sigma, split = split)
proc.time() - ptm

source("./functions/cbindTModels.R") # bind traffic models together ready for plotting

fcasts.all <- cbindTModels(fcasts,t,split,m)
fcasts.all <- fcasts.all %>% filter(link != "X1384" & link != "X2007" & link != "X2140" & link != "X2301") # exclude historic links

# tidy for plotting
fcasts.g <- gather(fcasts.all, Values, ujt, -dtime:-link)

#######################
# plot links 1:9
p <- ggplot(data = filter(fcasts.g, link %in% c("X425", "X432","X448", "X463", "X474", "X524", "X1025", "X1419", "X1447")), aes(x = dtime, y = ujt, group = Values, colour = Values))

p + geom_line(alpha=0.7) + #colour="grey60"
  scale_x_continuous(limits= c(6,21), breaks = seq(from = 6, to = 21, by = 3), labels=as.character(seq(6,21,3))) +
  facet_grid(link ~ date) +
  xlab("Time of day (24hr clock)") + ylab("Unit Journey Time (seconds/metre)") +
  scale_colour_manual(values = c("indianred1", "grey20"))
#ggsave(filename = "./outputs/Link_predictions_1-9.pdf",width = 15, height = 9)
ggsave(filename = "./outputs/Link_predictions_1-9.png",width = 220, height = 250, units="mm")


#######################
# plot links 10:18
p <- ggplot(data = filter(fcasts.g, link %in% c("X1592", "X1593", "X1616", "X1623", "X2052", "X2055", "X2079", "X2085", "X2324")), aes(x = dtime, y = ujt, group = Values, colour = Values))

p + geom_line(alpha=0.7) + #colour="grey60"
  scale_x_continuous(limits= c(6,21), breaks = seq(from = 6, to = 21, by = 3), labels=as.character(seq(6,21,3))) +
  facet_grid(link ~ date) +
  xlab("Time of day (24hr clock)") + ylab("Unit Journey Time (seconds/metre)") +
  scale_colour_manual(values = c("indianred1", "grey20"))
#ggsave(filename = "./outputs/Link_predictions_10-18.pdf",width = 15, height = 9)
ggsave(filename = "./outputs/Link_predictions_10-18.png",width = 220, height = 250, units="mm")
```

Find diagnostics from models
```{r Find diagnostics from models}

source("./functions/trafficDiagnostics.R") # function to collect diagnostics

fcast.info <- tInfo(fcast = fcasts) 

```


```{r Clean up}

rm(split, C, epsilon, sigma, m, p, t.fcasts)

save(list=ls(pattern="v\\.*day"), file="./data/SVR_allDay_training.RData")

```

