---
title: "SVM tuning AM"
author: "Mark Ruddy"
date: "20 April 2015"
output: html_document
---

# Tuning SVM model of traffic data in morning peaks
1. Get AM data from wednesdays
2. Train models using 1st order spatial weights
3. Try different lengths of hsitorical data for regression (m=c(3, 6, 12): 15, 30 and 60 minutes)
4. Choose two links for model assessment (X448, X1593)
5. Train with 80% training data

## Setup: data, functions and packages

```{r Setup data, functions and packages}
library(kernlab)
library(ggplot2)
library(grid)
library(dplyr)
library(tidyr)

source("./functions/Kernel_methods_lib.R")
source("./functions/trainModel.R") # train models with mutiple parameters
source("./functions/trainTraffic.R") # train models with mutiple parameters from multiple datasets

# datasets including spatial weights and Wednesday UJTs
load("./data/STDM_base.RData")

# rmse function
rmse <- function(obs, pred, na.rm=TRUE) {
  sqrt(mean((obs-pred)^2, na.rm=na.rm))
  }

# Wednesday UJTs
t <- ujt.repl.hist
```


Prepare data
```{r Prepare data}
# weight matrices as.matrix
w1.m <- as.matrix(w1[,-1])
w2.m <- as.matrix(w2[,-1])
w3.m <- as.matrix(w3[,-1])

# AM UJT data (<10am)

t.am <- t %>%
  filter(dtime<10)
```

## Train AM UJT for links X448 and X1593
```{r Train AM UJT}
# parmaters
c <- 2^rep(-5:4) # 
sig <- 2^rep(-9:2) # 
eps <-c(0.01, 0.1, 0.5) # 

# split at 960 is 20 days of AMs (20 days * 48 5min intervals)
v.am.m3 <- trainTraffic(ujt = select(t.am, -dtime:-date), weights = w1.m, link = c(3, 12), m = 3, C = c, epsilon = eps, sigma = sig, split = 960)

v.am.m6 <- trainTraffic(ujt = select(t.am, -dtime:-date), weights = w1.m, link = c(3, 12), m = 6, C = c, epsilon = eps, sigma = sig, split = 960)

v.am.m12 <- trainTraffic(ujt = select(t.am, -dtime:-date), weights = w1.m, link = c(3, 12), m = 12, C = c, epsilon = eps, sigma = sig, split = 960)

```

## Plot errors for X448
Combine x448 training for all m
```{r Combine x448 training for all m}
v3 <- v.am.m3$X448$error_mat
v6 <- v.am.m6$X448$error_mat
v12 <- v.am.m12$X448$error_mat

v448am <- rbind(v3, v6, v12)
v448am <- cbind(v448am, c(rep(3,nrow(v3)), rep(6,nrow(v6)), rep(12,nrow(v12))))
colnames(v448am)[8] <- "m" 
v448am <- as.data.frame(v448am)

rm(v3, v6, v12)
```

Plot multi-facet of X448 for training error ~ different epsilon and m
```{r Plot multi-facet of X448 training}

p <- ggplot(v448am, aes(x=factor(Sigma), y=factor(C), fill=TrainingError))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR training error - link X448 AM with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="Training RMSE"), low="#000066", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
ggsave(filename = "./outputs/TrainError_X448_AM.pdf",width = 15, height = 9)


```


Plot multi-facet of X448 for testing error - different epsilon and m
```{r Plot multi-facet of X448 testing}

p <- ggplot(v448am, aes(x=factor(Sigma), y=factor(C), fill=RMSE))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR test error - link X448 AM with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="Forecast RMSE"), low="darkred", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
ggsave(filename = "./outputs/TestError_X448_AM.pdf",width = 15, height = 9)

```



## Plot errors for X1593
Combine X1593 training for all m
```{r Combine X1593 training for all m}
v3 <- v.am.m3$X1593$error_mat
v6 <- v.am.m6$X1593$error_mat
v12 <- v.am.m12$X1593$error_mat

v1593am <- rbind(v3, v6, v12)
v1593am <- cbind(v1593am, c(rep(3,nrow(v3)), rep(6,nrow(v6)), rep(12,nrow(v12))))
colnames(v1593am)[8] <- "m" 
v1593am <- as.data.frame(v1593am)

rm(v3, v6, v12)
```

Plot multi-facet of X1593 for training error ~ different epsilon and m
```{r Plot multi-facet of X1593 training}

p <- ggplot(v1593am, aes(x=factor(Sigma), y=factor(C), fill=TrainingError))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR training error - link X1593 AM with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="Training RMSE"), low="#000066", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
ggsave(filename = "./outputs/TrainError_X1593_AM.pdf",width = 15, height = 9)


```

Plot multi-facet of X1593 for testing error ~ different epsilon and m
```{r Plot multi-facet of X1593 testing}

p <- ggplot(v1593am, aes(x=factor(Sigma), y=factor(C), fill=RMSE))

p + geom_tile() +
  facet_wrap(m ~ Epsilon, ncol=3) +
  labs(x=expression(paste("Kernel bandwidth, ", sigma)), y="Error penalisation, C") +
  ggtitle("SVR test error - link X1593 AM with 1st order spatial weights") +
  scale_fill_continuous(guide=guide_colourbar(title="Forecast RMSE"), low="darkred", high="white") +
  theme(panel.margin=unit(1,"cm"), axis.text.x=element_text(vjust=0.5,  size=10)) +
  scale_y_discrete(breaks = c(2^rep(-5:4)), labels = c("2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3", "2^4")) +
  scale_x_discrete(breaks = c(2^rep(-10:3)), labels = c("2^-10", "2^-9", "2^-8", "2^-7", "2^-6", "2^-5", "2^-4", "2^-3", "2^-2", "2^-1", "1", "2^1", "2^2", "2^3"))
ggsave(filename = "./outputs/TestError_X1593_AM.pdf",width = 15, height = 9)

```

Detailed data checking
```{r Detailed data checking}

v448am %>% group_by(Epsilon, m) %>% summarise(min(TrainingError), min(RMSE)) %>% View()

v1593am %>% group_by(Epsilon, m) %>% summarise(min(TrainingError), min(RMSE)) %>% View()

```


```{r Clean up}

rm(v448am, v1593am)

save(list=ls(pattern="v\\.*am"), file="./data/SVR_am_training.RData")

```







